---
title: "global_climate_change"
author: "Ryan Whalen"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

```{r message = FALSE}
library(tidyverse)
library(lubridate)
library(modelr)
```

### Changes in Modern Temperatures
```{r}
#Import "Global-mean, monthly, seasonal, and annual means" data from https://data.giss.nasa.gov/gistemp/
global_means <- read_csv("data_raw/GLB.Ts+dSST.csv", skip = 1)

#Selected only the year and annual temp. data located under `Year` and ``J-D`` columns. 
#Changed the names of the the year and temperature columns to `year` and `annual_temp`
#The units of temperature are Celcius but are provided as the difference from the ean temperature between 1951-1980
global_temp_annual <- global_means %>% 
  select("Year", "J-D") %>%
  rename("year" = 1) %>% 
  rename("annual_temp" = 2) %>%
  filter(year < 2020)
global_temp_annual$annual_temp <- as.numeric(global_temp_annual$annual_temp)

#Scatter plot of modern temperature change over time
ggplot(data = global_temp_annual, mapping = aes(x = year, y = annual_temp)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Years",
       y = "Celcius (anomalies)") 

#Creates linear regression model
fit <- lm(formula = annual_temp ~ year, global_temp_annual)
summary(fit)
```

#### Changes in Modern Temperatures (Post 1950)
```{r}
#Adjusted the trend line so that it only looks at the most recent decades since many scientists claim that drastic changes in global temperature began around 1950 when personal transportation became common
#Dataset with data after 1950
post_1950 <- global_temp_annual %>% 
  filter(year > 1950)

#Scatter plot of modern temperature change over time starting at 1950  
ggplot(data = post_1950, mapping = aes(x = year, y = annual_temp)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Years",
       y = "Celcius (anomalies)") 

#Creates linear regression model to determine current rate of change for temperature data
post_1950_fit <- lm(formula = annual_temp ~ year, post_1950)
summary(post_1950_fit)
```

### Changes in Modern Atmospheric CO2 
```{r}
#Import annual mean atmospheric CO2 data downloaded from https://www.esrl.noaa.gov/gmd/ccgg/trends/data.html
co2_data <- read_table("data_raw/co2_annmean_mlo.txt", col_names = TRUE, skip = 56)

#Select the "year" and "mean" columns
co2_data <- co2_data %>% 
  select("year","mean")

#Graph of CO2 vs. Time 
gg_co2 <- ggplot(data = co2_data, mapping = aes(x = year, y = mean)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Years",
       y = "Mean CO2 Levels") 

#Creates linear regression model to determine current rate of change for atmospheric CO2 data
RoC_fit <- lm(mean ~ year, co2_data)
summary(RoC_fit)
```


#### Relationship between modern temperatures and atmospheric CO2
```{r}
#Combines temperature and CO2 data
temp_co2_data <- inner_join(co2_data, global_temp_annual)
view(temp_co2_data)

#Graph of relationship between atmospheric CO2 and temperature
ggplot(data = temp_co2_data, mapping = aes(x = mean, y = annual_temp)) +
  geom_point() +
  labs(x = "Mean CO2 Levels",
       y = "Annual Temperatures in Celcius (anomalies") 

#Creates linear regression model
CO2_temp_fit <- lm(annual_temp ~ mean, temp_co2_data)
summary(CO2_temp_fit)
```

### Pre-historic Temperature Data
```{r}
#Import Vostok ice core data from the Carbon Dioxide Information Analysis Center (CDIAC) 
#https://cdiac.ess-dive.lbl.gov/trends/temp/vostok/jouz_tem.htm
vostok_temp_data <- read_table("data_raw/vostok.1999.temp.dat.txt", col_names = FALSE, skip = 60)

#Renamed the columns
vostok_temp_data <- vostok_temp_data %>% 
  rename("Depth (m)" = 1) %>% 
  rename("Age of the ice (yr BP)" = 2) %>% 
  rename("Deuterium content of the ice (delta D)" = 3) %>% 
  rename("Temperature Variation (deg C)" = 4)

#Creates new variable and calculates the paleo-temperature at Vostok based on the following formula
#Temperature (in degrees C) = -55.5 + (delta D + 440) / 6
vostok_temp_data$Paleo_Temperature <- -55.5 + (vostok_temp_data$`Deuterium content of the ice (delta D)` + 440) / 6

#Renamed "Temperature Variation" column as "Paleo-Temperature (deg C)"
vostok_temp_data <- vostok_temp_data %>% 
  rename("Paleo-Temperature (deg C)" = 5)

#Graph of ice-cre temperature data using ice age as the independent variable
ggplot(data = vostok_temp_data, mapping = aes(x = vostok_temp_data$`Age of the ice (yr BP)`, y = vostok_temp_data$`Paleo-Temperature (deg C)`)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Age of the ice (yr BP)",
       y = "Paleo-Temperature (deg C)") 

#Creates linear regression model
vostok_fit <- lm(vostok_temp_data$`Paleo-Temperature (deg C)` ~ vostok_temp_data$`Age of the ice (yr BP)`, vostok_temp_data)
summary(vostok_fit)

#Filter dataset to period of time with fastest rate of change
vostok_fast_change <- vostok_temp_data %>% 
  filter(vostok_temp_data$`Age of the ice (yr BP)` > 25000 & vostok_temp_data$`Age of the ice (yr BP)` < 50000)

#Graph of new dataset
gg_change <- ggplot(data = vostok_fast_change, mapping = aes(x = vostok_fast_change$`Age of the ice (yr BP)`, y = vostok_fast_change$`Paleo-Temperature (deg C)`)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Age of the ice (yr BP)",
       y = "Paleo-Temperature (deg C)") 
gg_change

#Creates a linear regression model
vostok_change_fit <- lm(vostok_fast_change$`Paleo-Temperature (deg C)` ~ vostok_fast_change$`Age of the ice (yr BP)`, vostok_fast_change)
summary(vostok_change_fit)
```
### Pre-historic CO2 Data
```{r} 
#Import Vostok ice core CO2 data from CDIAC website
#https://cdiac.ess-dive.lbl.gov/trends/co2/vostok.html
vostok_co2_data <- read_delim("data_raw/vostok.icecore.co2.txt", delim = "\t", col_names = FALSE, skip = 20)
vostok_co2_data <- vostok_co2_data %>% 
  rename("Depth (m)" = 1) %>% 
  rename("Age of the ice (yr BP)" = 2) %>% 
  rename("Mean age of the air (yr BP)" = 3) %>% 
  rename("CO2 Concentration (ppmv)" = 4)

#Graph of CO2 concentration as a funtion of (gas) age using full time period
gg_co2_conc <- ggplot(data = vostok_co2_data, mapping = aes(x = `Mean age of the air (yr BP)`, y = `CO2 Concentration (ppmv)`)) +
  geom_point() +
  labs(x = "Mean age of the air (yr BP)",
       y = "CO2 Concentration (ppmv)")

#New dataset for fastest rate of change for CO2 in pre-historic time period
vostok_co2_focus_data <- vostok_co2_data %>% 
  filter(vostok_co2_data$`Age of the ice (yr BP)` > 128000 & vostok_co2_data$`Age of the ice (yr BP)` < 145000)

#Graphs the new dataset
ggplot(data = vostok_co2_focus_data, mapping = aes(x = `Mean age of the air (yr BP)`, y = `CO2 Concentration (ppmv)`)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Mean age of the air (yr BP)",
       y = "CO2 Concentration (ppmv)")

#Creates a linear regression model
focus_fit <- lm(vostok_co2_focus_data$`CO2 Concentration (ppmv)` ~ vostok_co2_focus_data$`Mean age of the air (yr BP)`, vostok_co2_focus_data)
summary(focus_fit)
```

#### Comparison of modern and pre-historic rate of changes in CO2
```{r}
#Graph of fastest natural rate of change in CO2 from pre-historic data
gg_co2_conc
#Graph of modern rate of change in CO2
gg_co2
```
### Hypothesis 
**Question 11:** How do current (i.e., since 1950) changes in 
atmospheric CO2 concentration and average global temperature compare to 
pre-historic (i.e., in the past hundreds of thousands of years) changes in these
variables?

**Answer 11:**
Current changes in atmospheric CO2 and temperature are increasing at a steady rate while the CO2 atmospheric concentration and temperature changes in the pre-historic era fluctuated greatly and weren't steadily increasing
**Question 12:** What does this suggest about whether recent changes in 
temperature are due to natural or anthropogenic (human) factors? 

**Answer 12:**
This highly suggests that the recent changes in temperature are due to anthropogenic factors
**Question 13:** Is plausible that recent increase in atmospheric carbon 
dioxide is a result of natural fluctuations and not human-induced?

**Answer 13:**
It could be natural fluctuations but it likely isn't. The CO2 naturally flucated in pre-historic times but never really went above 300ppmv of CO2 while in todays world we have over 400ppmv of atmospheric CO2
## Citation

This module was initially developed as an Excel-based exercise by:

O’Reilly, C.M., D.C. Richardson, and R.D. Gougis. 15 March 2017. Project EDDIE: 
Climate Change. Project EDDIE Module 8, Version 1. 
http://cemast.illinoisstate.edu/data-for-students/modules/climate-change.shtml. 
Module development was supported by NSF DEB 1245707

